Relationships in the extremes and their influence on competition and coexistence

### 2021-07-06 paper/Paper.Rnw
further progress in writing sup mat sections associated with Methods

%Jasmin, I think the below makes a pretty good start at tryinbg to nail down a good way to express our goals
%in the last para of the Intro, so please incorporate these into your last para. If you have a better way
%to phrase these, please put both so we can compare at some point when we meet.
G1) To augment the modern theory of coexistence (MTC) to show how asymmetric tail association (ATA)
plays a role. The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. We here examine mechanisms of coexistence that apply in 
models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore using a simple model the relative magnitudes of the contributions to coexistence that come
from ATA compared to other mechanisms.

@@

For the lottery model, when $i$ is an invader, the competition experienced by either 
species is obtained from the expression for $C_i=C_j$ by setting $N_i=0$ and $N_j=N$; and likewise to get
$\olr_{i \bs i}$ from the expression for $r_i$ for the model. Thus 
$C_{i \bs i}(t)=C_{j \bs i}(t)=B_j(t)/\delta=E_j(t)/\delta$, and 
$\olr_{i \bs i} = \E[\ln(1-\delta+\delta B_i/B_j)]=\E[\ln(1-\delta+E_i /C_{i \bs i})]$. 
Likewise, 
$\olr_{j \bs i} = \E[\ln(1-\delta+\delta B_j/B_j)]=\E[\ln(1-\delta+E_j/C_{j \bs i})]$,
though we already knew this equals $0$. 
An expression for $\olrs_{i \bs i}$ is obtained by replacing $(E_i,C_{i \bs i})=(B_i,B_j/\delta)$
by a random variable with the same marginals, but independent components, $(B_i^\shp,B_j^\shp /\delta)$.
Thus $\olrs_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\shp/B_j^\shp)]$. Likewise,
$\olrs_{j \bs i}$ is computed by replacing $(E_j,C_{j \bs i})$ in an analogous way; we use 
$(E_j^\shp,C_{j \bs i}^\shp)=(B_j^\shp, B_j /\delta)$, so 
$\olrs_{j \bs i}=\E[\ln(1-\delta+\delta B_j^\shp / B_j)]$.
We show in SI section \ref{SI-SIsect:scaling_factors}
that $q_{ij}=\overline{B}_i/((1-\delta)\overline{B}_j+\delta \overline{B}_i)$. 
%***DAN: Note this might be wrong. Some of the language in Ellner et al (2016) indicates q_{ij} here should be 1,
%even if \mu_1 and \mu_2 are distinct, as long as we use the same \delta for both species (which we do).
%But when I try to follow the definitions of the same paper to compute q_{ij} I get the quantity listed above,
%which is not 1. I'll need to get to the bottom of this, possibly by speaking to Ellner, and then put the right 
%value here, and also write the corresponding sup mat section accordingly (though you may not need that section, 
%it may be enough to just cite Ellner and Chesson if you go with q_ij=1.
Since we assumed log-normally distributed $B_i$, 
$\overline{B}_i=\exp(\mu_i+\sigma^2/2)$, so $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$.
%***DAN: This statement also needs adjusting if you go with q_ij=1
Thus $\Delta_i^{(EC)}$ for the lottery model can be computed by using simulations \citep{Ellner_2016}
to very accurately approximate $\olr_{i \bs i}$, $\olrs_{i \bs i}$,
and $\olrs_{j \bs i}$, and then plugging them and $q_{ij}$ into (\ref{eq:storage_effects_gen}).

We now specify how to decompose storage effects into
contributions due to asymmetric tail associations and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between invader and resident mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the invader/resident growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[E \nat C]} + \Delta_i^{[EC]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to 
asymmetries of tail association between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATA).
The term $\Delta_i^{[EC]}$ is the component of storage effects due to $EC$ correlation \emph{per se}, remaining after 
having removed the effects of asymmetric tail associations.
%The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
%this is a decompostion of storage effects, which were denoted by \cite{Ellner_2019} using a superscript $(EC)$,
%i.e., $\Delta_i^{(EC)}$, to indicate that storage effects are, in turn, part of a decomposition of a difference between invader and
%resident growth rates (see below).  
The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decompostion within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term within the decomposition provided by the MTC \citep{Ellner_2019} -- see below for details.

@@

%Goals, for reference:
%G1) To augment the modern theory of coexistence (MTC) to show how asymmetric tail association (ATA)
%plays a role. The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
%We address how ATA contributes to these components. We here examine mechanisms of coexistence that apply in 
%models exhibiting solely temporal variation; spatial variation can be considered in future work.
%G2) To explore using a simple model the relative magnitudes of the contributions to coexistence that come
%from ATA compared to other mechanisms.

@@

Left- or right-tail associated noise $(b_1,b_2)$, with $b_i$ normally distributed with mean $\mu_i$ and standard deviation 
$\sigma$, was generated as pictured in Fig. \ref{fig:noise}, with details in SI section \ref{SI-sect:noise}.
To achieve goal G2 from the Introduction, i.e., to explore using a simple model the relative magnitudes of the contributions 
to coexistence that come from ATA versus other mechanisms, we plotted the terms $\Delta_i^0$, $\Delta_i^E$, $\Delta_i^C$, 
$\Delta_i^{(E\shp C)}$, $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ [see (\ref{eq:full_decomp})] for the lottery model
against the independent parameters of the model, $\mu_1$, $\mu_2$, $\sigma$ and $\delta$; and then we compared their
values on the plots. Some mathematical shortcuts, summarized here, greatly simplified this plotting. First, we showed that, for the 
lottery model, none of the components in (\ref{eq:full_decomp}) depends independently on $\mu_1$ and $\mu_2$,
instead they all depend only on the difference $\mu_1-\mu_2$ (SI section \ref{SI-sect:math_simplifications}); so we only plotted 
against $\mu_1 - \mu_2$. Second, we showed that the values of all components in (\ref{eq:full_decomp}) were the same 
for both our left- and right-tail associated noises (SI section \ref{SI-sect:math_simplifications}); so we only generated plots 
for left-tail associated noise. For each combination of $\delta=W, X, Y, Z$
%***DAN: Jasmin, please select 4 sensible values, including \delta=1, and put the ones you select here
and $\mu_1-\mu_2=A,B,C,D$,
%***DAN: Jasmin, please select 3-4 sensible values and put the ones you select here
we plotted all the components of (\ref{eq:full_decomp}) against $\sigma$. Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully large for the lottery model.
%***DAN: Jasmin, if you select 4 values for mu1-mu2 and 4 for delta, then Fig. 3 is going to have 16 panels arranged in a 
%4 by 4 grid, each of which will plot all the components against sigma. You'll need to use different colors and/or line
%types for the components and have a legend.
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).
Computationally efficient means of carrying out these simulations, for the lottery model, are elaborated in 
SI section X.
%***DAN: Dan to add the reference to the right sup mat section once things get sufficiently organized in sup mat to know what section.

>>>>

%Jasmin, I think the below makes a pretty good start at tryinbg to nail down a good way to express our goals
%in the last para of the Intro, so please incorporate these into your last para. If you have a better way
%to phrase these, please put both so we can compare at some point when we meet.
G1) To augment the modern theory of coexistence (MTC) to show how asymmetric tail association (ATA)
plays a role. The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. We here examine mechanisms of coexistence that apply in 
models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATA compared to other mechanisms.

@@

For the lottery model, when $i$ is an invader, the competition experienced by either 
species is obtained from the expression for $C_i=C_j$ by setting $N_i=0$ and $N_j=N$; and likewise to get
$\olr_{i \bs i}$ from the expression for $r_i$ for the model. Thus 
$C_{i \bs i}(t)=C_{j \bs i}(t)=B_j(t)/\delta=E_j(t)/\delta$, and 
$\olr_{i \bs i} = \E[\ln(1-\delta+\delta B_i/B_j)]=\E[\ln(1-\delta+E_i /C_{i \bs i})]$. 
Likewise, 
$\olr_{j \bs i} = \E[\ln(1-\delta+\delta B_j/B_j)]=\E[\ln(1-\delta+E_j/C_{j \bs i})]$,
though we already knew this equals $0$. 
An expression for $\olrs_{i \bs i}$ is obtained by replacing $(E_i,C_{i \bs i})=(B_i,B_j/\delta)$
by a random variable with the same marginals, but independent components, $(B_i^\shp,B_j^\shp /\delta)$.
Thus $\olrs_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\shp/B_j^\shp)]$. Likewise,
$\olrs_{j \bs i}$ is computed by replacing $(E_j,C_{j \bs i})$ in an analogous way; we use 
$(E_j^\shp,C_{j \bs i}^\shp)=(B_j^\shp, B_j /\delta)$, so 
$\olrs_{j \bs i}=\E[\ln(1-\delta+\delta B_j^\shp / B_j)]$.
We show in SI section \ref{SI-SIsect:scaling_factors}
that $q_{ij}=\overline{B}_i/((1-\delta)\overline{B}_j+\delta \overline{B}_i)$. 
Since we assumed log-normally distributed $B_i$, 
$\overline{B}_i=\exp(\mu_i+\sigma^2/2)$, so $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$.
We alternatively also consider $q_{ij}=1$, for reasons explained in SI section \ref{SI-SIsect:scaling_factors}.
%Say this here if it turns out to be true, or say something like it:
%These alternatives turn out to make very little difference to our results. 
Thus $\Delta_i^{(EC)}$ for the lottery model can be computed by using simulations \citep{Ellner_2016}
to very accurately approximate $\olr_{i \bs i}$, $\olrs_{i \bs i}$,
and $\olrs_{j \bs i}$, and then plugging them and $q_{ij}$ into (\ref{eq:storage_effects_gen}).

We now specify how to decompose storage effects into
contributions due to asymmetric tail associations and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between invader and resident mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the invader/resident growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[E \nat C]} + \Delta_i^{[EC]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to 
asymmetries of tail association between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATA).
The new symbol should correspondingly be pronounced ``partial sharp''.
The term $\Delta_i^{[EC]}$ is the component of storage effects due to $EC$ correlation \emph{per se}, remaining after 
having removed the effects of asymmetric tail associations.
%The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
%this is a decompostion of storage effects, which were denoted by \cite{Ellner_2019} using a superscript $(EC)$,
%i.e., $\Delta_i^{(EC)}$, to indicate that storage effects are, in turn, part of a decomposition of a difference between invader and
%resident growth rates (see below).  
The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decompostion within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term within the decomposition provided by the MTC \citep{Ellner_2019} -- see below for details.

@@

%Goals, for reference:
%G1) To augment the modern theory of coexistence (MTC) to show how asymmetric tail association (ATA)
%plays a role. The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
%We address how ATA contributes to these components. We here examine mechanisms of coexistence that apply in 
%models exhibiting solely temporal variation; spatial variation can be considered in future work.
%G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
%from ATA compared to other mechanisms.

@@

Left- or right-tail associated noise $(b_1,b_2)$, with $b_i$ normally distributed with mean $\mu_i$ and standard deviation 
$\sigma$, was generated as pictured in Fig. \ref{fig:noise}, with details in SI section \ref{SI-sect:noise}.
To achieve goal G2 from the Introduction, i.e., to explore using a simple model the relative magnitudes of the contributions 
to coexistence that come from ATA versus other mechanisms, we plotted the terms $\Delta_i^0$, $\Delta_i^E$, $\Delta_i^C$, 
$\Delta_i^{(E\shp C)}$, $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ [see (\ref{eq:full_decomp})] for the lottery model
against the independent parameters of the model, $\mu_1$, $\mu_2$, $\sigma$ and $\delta$; and then we compared their
values on the plots. Some mathematical shortcuts, summarized here, greatly simplified this plotting. First, we showed that, for the 
lottery model, none of the components in (\ref{eq:full_decomp}) depends independently on $\mu_1$ and $\mu_2$;
instead they all depend only on the difference $\mu_1-\mu_2$ (SI section \ref{SI-sect:math_simplifications}). So we only plotted 
against $\mu_1 - \mu_2$. Second, we showed that the values of all components in (\ref{eq:full_decomp}) were the same 
for both our left- and right-tail associated noises (SI section \ref{SI-sect:math_simplifications}). So we only generated plots 
for left-tail associated noise. For each combination of $\delta=W, X, Y, Z$
%***DAN: Jasmin, please select 4 sensible values, including \delta=1, and put the ones you select here
and $\mu_1-\mu_2=A,B,C,D$,
%***DAN: Jasmin, please select 3-4 sensible values and put the ones you select here
we plotted all the components of (\ref{eq:full_decomp}) against $\sigma$. Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully large for the lottery model, compared to 
the contributions of other mechanisms.
%***DAN: Jasmin, if you select 4 values for mu1-mu2 and 4 for delta, then Fig. 3 is going to have 16 panels arranged in a 
%4 by 4 grid, each of which will plot all the components against sigma. You'll need to use different colors and/or line
%types for the components and have a legend.
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).
Computationally efficient means of carrying out these simulations, for the lottery model, are elaborated in 
SI section \ref{SI-sect:efficient_computation}.
%***DAN: Dan, you still need to write that sup mat section
